---
title: "Seeing Groups, Not Just Gradients:"
subtitle: "Supervised Class Differentiation to Better Interpret Beta Diversity"
author: "<br>Nicholas Spyrison, PhD<br>spyrison@gmail.com"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

```{r, eval=TRUE, echo=FALSE}
## I like lucy theme, but tabels and figures are borked somethign bad.
knitr::opts_chunk$set(
  echo = FALSE, ## Code
  include = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  out.height = "70%",
  fig.align = "center", fig.dpi = 300, fig.retina = 3,
  cache = FALSE, cache.lazy = FALSE
)
```

```{r}
## Packages
suppressPackageStartupMessages({suppressWarnings({
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(xaringanthemer)
  # library(phyloseq)
  # library(vegan)
})})

xaringanthemer::style_mono_dark(base_color = "#cbf7ed")
options(tidyverse.quiet = TRUE)
theme_set(theme_bw())
here::i_am("BetaSpaces_Presentation.Rmd")

## Load artifacts
.verb <- ifelse(interactive(), TRUE, FALSE)
load(file = here::here(
  "./zScratchpad/ACVD_Inputs/outputs/ns_diversity_spaces.rda"), verbose = .verb)
```

# Terminology

- Mixed audiences; write Biology and interpret to Statistics

```{r, echo=FALSE}
tribble(
  ~Term,             ~Alias,                     ~Notaion_dimension,
  "Sample",          "Site, Observation, Row",   "n",
  "Variable",        "Colunm, feature, Metric",  "p", # "Dimension",
  "Abundance table", "Data",                     "n x p",
  "Metadata",        "Context data",             "n rows",
  "Distance*",        "Dissimilarity",            "n x n', where n' <= n*",
  "Ordination",      "Embedding, reduced scape", "n x d, where d <= n' <= n"
) %>%
  knitr::kable() %>%
  kableExtra::footnote(
    general = "* Biological distances sometimes don't meet the triangle inequality, leading to negative some eigenvalues. By default, most implementations drop negative eigenvalued components rather than diluting the others by adding the difference to all eigenvalues.",
    #general_title = "", 
    fixed_small_size = TRUE)
```

---

# Ecological Taxonomic Diversity

- *Alpha diversity, α (1),* richness and evenness within a site/sample
    - Typically choose 2-3 of 7 metrics, **visualized independently**
- **Beta diversity, β (1),** compositional differences between sites/samples
    - Typically *10s-1000s of components* (dissimilarity of taxa abundance), **visualized as 2 components of 1 or 2 ordination method**

.footnote[
(1) R. H. Whittaker (1960) Vegetation of the Siskiyou Mountains, Oregon and California. Ecological Monographs, 30, 279–338. doi:10.2307/1943563
<br><br><br>
]

---

# Alpha Diversity, α

- Richness describes the unique number of taxa
- Evenness describes the distribution of taxa abundances
- Diversity is a measure of both richness and evenness

```{r, echo=FALSE}
tribble(
  ~Metric,      ~Class,      ~Context,  ~Description,
  "Observed",   "Richness",  "Wide",    "Count of observed taxa. Sensitive to rare taxa",
  "Shannon",    "Diversity", "Wide",    "Eveness and richness, log-based, range [0, log(n_tax)]",
  "Simpson",    "Evenness",  "Wide",    "Prob two randomly selected individuals are different species, range [0, 1]",
  "InvSimpson", "Evenness",  "+1",      "Inverse of Simpson (1/D). More intuitive scale",
  "Chao1",      "Richness",  "Non-ASV", "Accounting for unseen rare species, better for undersampled data",
  "ACE",        "Richness",  "Non-ASV", "Abundance-based Chao estimator, weights by abundance",
  "Fisher",     "Diversity", "?",       "Parametric index balancing richness and evenness, assumes log-series") %>% knitr::kable()
```

---

# The Data

Follow the gut microbiome analysis in Jie et al, 2017 (3). Observed 385 samples of the gut microbiome in individuals with Artherosclerotic CardioVascular Disease (ACVD, "heart disease") compared to healthy individuals in a Chinese cohort. For our purpose, **goal is to visualize for taxa associated differntially by heart disease.**

- Data (2): 370 obs x 2397 taxa (relative abundances) -- `otu_table(physeq)`
- Metadata (2): 370 obs x 9 variables (Heart Disease, Sex, Age, Triglycerides, Cholesterols) -- `sample_data(physeq)`
- Alpha diversity: 370 obs x 4 within-site measures of richness, evenness, and diversity -- `estimate_richness(physeq, measures = c("Observed", "Shannon", "InvSimpson", "Fisher")`
- Beta diversity, PCoA on Bray-Curtis distance: TODO
- Beta diversity, CAP on Bray-Curtis distance: TODO

.footnote[
(2) Jie, Z., Xia, H., Zhong, S. L., Feng, Q., Li, S., Liang, S., ... & Kristiansen, K. (2017). The gut microbiome in atherosclerotic cardiovascular disease. Nature communications, 8(1), 845.
]

---

# Alpha Space

.pull-left[
Typically 2 to 7 alpha diversity metrics, visualized like...

````
estimate_richness(...) %>%
  as_tibble(rownames = "SampleID") %>%
  left_join(metadata) %>%
  pivot_longer() %>%
  ggplot() + ...
````

Apply an appropriate statistical test, *eg.* T, Wilcoxon, Kurskal-Wallis, or Baysian variants
]

.pull-right[
```{r, echo=F, fig.dpi=300}
# viridis::viridis(2, option = "A", begin = 0.3, end = 0.7) %>% dput()
acvd_colors <- c("#641A80FF", "#F76F5CFF")
ggplot(jie_alpha_long) +
  geom_boxplot(aes(x = ACVD, y = value, fill = ACVD)) +
  ylab('Alpha Diversity') +
  scale_fill_manual(values = acvd_colors) +
  facet_wrap(~Diversity_Measure, scales = 'free') +
  theme_bw()
```
]

---

# Alpha Space 2

.pull-left[
## Recomendations

1. Consider `geom_violin() + geom_jitter()` instead of `geom_boxplot()` to 
1. Apply color-blind friendly palettes with `scale_color_brewer()` and `scale_fill_brewer()` and [ColorBrewer2.org](https://colorbrewer2.org/).
1. Consider adding arrows to indicate domain preferred domain directions, especially for Simpson's Index (raw) where lower is more diverse
]

.pull-right[
```{r}
y_ranges <- jie_alpha_long |>
  group_by(Diversity_Measure) |>
  summarise(y_min = min(value, na.rm = TRUE),
            y_max = max(value, na.rm = TRUE),
            .groups = "drop") |>
  mutate(arrow_y_start = y_min + 0.05 * (y_max - y_min),
         arrow_y_end = y_max - 0.05 * (y_max - y_min))

ggplot(jie_alpha_long) +
  geom_violin(aes(x = ACVD, y = value, fill = ACVD), alpha = .3) +
  geom_jitter(aes(x = ACVD, y = value, color = ACVD)) +
  ## Arrows
  geom_segment(
    data = y_ranges,
    aes(x = .5, xend = .5, y = arrow_y_start, yend = arrow_y_end),
    arrow = arrow(ends = "last", length = unit(0.3, "cm"), type = "closed"),
    color = "green4", size = 1.1, inherit.aes = FALSE) +
  ylab("Alpha Diversity") +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~Diversity_Measure, scales = "free") +
  theme_bw()
```
]

---

# Alpha Space 3

.pull-left[
## Recomendations

- More information when pairing metrics:
    - **Visualize where evenness and richness disagree**
    - More take-aways than the last two figures
    - Scatterplot matrix(3) (aka SPLOM, pairs plot), `GGally::ggpairs()`
    - 2d Scatter plot with tooltips on sample/site & metadata with `plotly::ggplotly()`

.footnote[
(3) Hartigan, J. A. (1975). Clustering algorithms.
<br>John Wiley & Sons, Inc..
]
]

.pull-right[
```{r}
GGally::ggpairs(jie_alpha_wide, columns = 2:5,
                aes(color = ACVD, fill = ACVD, shape = ACVD)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2")
```
]

---

# Ordination, Linear

- After converting abundance table to a distance there is a large resulting space

```{r}
tribble(
  ~Goal, ~Data_Method, ~Distance_Method, ~Year_diff,  ~Supervision,
  "Maximize variance explained", "PCA, 1901", "PCoA, 1966", "65 Years", "Unsupervised",
  "Maximize group separation", "LDA, 1936", "CAP, 2003", "67 Years", "Supervised",
) %>% knitr::kable()
```

- Principal Component Analysis (PCA, 5, 1901)
- Principal Coordinate Analysis (PCoA, 6, 1966, aka cMDS, NMDS, PO)
- Linear Discriminant Analysis (LDA, 7, 1936)
- Canonical Analysis of Principal coordinates (CAP, 8, 2003)
- *PCA is to PCoA as LDA is to CAP*

.footnote[
(4) Pearson, K. (1901). "On Lines and Planes of Closest Fit to Systems of Points in Space." Philosophical Magazine, 2(11), 559–572. (The initial formal description).
<br>(5) Gower, J. C. (1966). "Some distance properties of latent root and vector methods used in multivariate analysis." Biometrika, 53, 325–338.
<br>(6) Fisher, R. A. (1936). "The use of multiple measurements in taxonomic problems." Annals of Eugenics, 7(2), 179–188.
<br>(7) Anderson, M. J., & Willis, T. J. (2003). "Canonical analysis of principal coordinates: a useful method of constrained ordination for ecology." Ecology, 84(3), 511–525.
]

---

# Beta Diversity -- PCoA on Bray-Curtis 

.pull-left[
## Typically

- Viewed as PC1:PC2
    - Implicitly discard 80% variance in the other 200 components
- Discussion on separability or unique regions of taxa on a known group
    - **PCoA orients directions of variance, not group separation; let's try CAP**
]

.pull-right[
```{r}
(pcoa2 <- jie_bray_pcoa_embed %>%
  ggplot(aes(x = `PC1 (12%)`, y = `PC2 (7.3%)`, color = ACVD)) +
  geom_point(aes(shape = ACVD), size = 2) +
  stat_ellipse(level = 0.95, linewidth = 1.1) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "PCoA1:2, Bray-Curtis Distance",
       color = "Heart\nDisease",
       shape = "Heart\nDisease"))

pcoa4 <- jie_bray_pcoa_embed %>%
  GGally::ggpairs(columns = 1:4,
                  aes(color = ACVD, fill = ACVD, shape = ACVD)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "PCoA1:4, Bray-Curtis Distance")
```
]

---

# Beta Diversity -- CAP on Bray-Curtis 

.pull-left[
- `k-1` CAP components; x-axis maximizes group separation
- MDS (aka PCoA, on the remaining space); y-axis maximizes variation (of the remaining space)
- Could display percent variance, but not the selection metric, and will be lower than PCoA
- However, we're still ignoring 200 components; let's try scatterplot martices
]

.pull-right[
```{r}
(cap2 <- jie_bray_cap_embed %>%
  ggplot(aes(x = CAP1, y = MDS1, color = ACVD)) +
  geom_point(aes(shape = ACVD), size = 2) +
  stat_ellipse(level = 0.95, linewidth = 1.1) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "CAP1:2, Bray-Curtis Distance",
       color = "Heart\nDisease",
       shape = "Heart\nDisease"))
```
]

---

# Beta Diversity -- SPLOM CAP1:4 on Bray-Curtis 

.pull-left[
- The number of components to embed or visualize is left as an exercise for the annalist
- Marginal density validates that CAP1 separates groups best
- Including MDS2:MDS3 shows views with outlying samples that would be missing in 2D
]

.pull-right[
```{r}
(cap4 <- jie_bray_cap_embed %>%
  GGally::ggpairs(columns = 1:4,
                  aes(color = ACVD, fill = ACVD, shape = ACVD)) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(title = "CAP1:4, Bray-Curtis Distance"))
```
]

---

# Beta Diversity -- PCoA vs CAP in 2 components

.pull-left[
```{r}
pcoa2 + theme(legend.position = "bottom")
```
]
.pull-right[
```{r}
cap2 + theme(legend.position = "bottom")
```
]

---

# Beta Diversity -- PCoA vs CAP, 4

.pull-left[
```{r}
pcoa4
```
]
.pull-right[
```{r}
cap4
```
]

---

# Summary

## Alpha Diversity

- Try `geom_violin()` instead of `geom_boxplot()`, regardless, use `geom_jitter()`
- View metrics as 2D scatterplot or scatterplot matrix for more information

## Beta Diversity

- Try CAP instead of PCoA, orders for group differences, which is what is dicussed
- View metrics as 2D scatterplot or scatterplot matrix for more information

---

# Citations -- Methods and Data 

- (1) R. H. Whittaker (1960) Vegetation of the Siskiyou Mountains, Oregon and California. Ecological Monographs, 30, 279–338. doi:10.2307/1943563
- (2) Jie, Z., Xia, H., Zhong, S. L., Feng, Q., Li, S., Liang, S., ... & Kristiansen, K. (2017). The gut microbiome in atherosclerotic cardiovascular disease. Nature communications, 8(1), 845.
- (3) Hartigan, J. A. (1975). Clustering algorithms. John Wiley & Sons, Inc..
- (4) Pearson, K. (1901). "On Lines and Planes of Closest Fit to Systems of Points in Space." Philosophical Magazine, 2(11), 559–572. (The initial formal description).
- (5) Gower, J. C. (1966). "Some distance properties of latent root and vector methods used in multivariate analysis." Biometrika, 53, 325–338.
- (6) Fisher, R. A. (1936). "The use of multiple measurements in taxonomic problems." Annals of Eugenics, 7(2), 179–188.
- (7) Anderson, M. J., & Willis, T. J. (2003). "Canonical analysis of principal coordinates: a useful method of constrained ordination for ecology." Ecology, 84(3), 511–525.

---

# Citations -- R Packages

- `phyloseq`: McMurdie, Stacy, and Steven Holmes. "Discovery and Characterization of the Human Microbiome through High-Throughput 16S rRNA Gene Sequencing." *PLoS ONE*, vol. 8, no. 4, 2013, e61217.
- `vegan`: Oksanen, Jari, et al. *vegan: Community Ecology Package*. 2026.
- `GGally`: Schloerke, Claus, et al. *GGally: Extension to 'ggplot2'*. 2025.
- `plotly`: Sievert, Carson. *Interactive Web-Based Data Visualization with R, plotly, and shiny*. 2020.
- `ggplot2`: Wickham, Hadley. *ggplot2: Elegant Graphics for Data Analysis*. 2016.
- `dplyr`: Wickham, Hadley, et al. *dplyr: A Grammar of Data Manipulation*. 2026.
- `knitr`: Xie, Yihui. *knitr: A General-Purpose Package for Dynamic Report Generation in R*. 2025.
- `R`: R Core Team. *R: A Language and Environment for Statistical Computing*. 2026.

---
class: center, inverse

# Thank you

## Questions

Nicholas Spyrison, PhD
<br>spyrison@gmail.com

---
class: center, inverse

# Appendix

---

# Ecological Taxonomic Diversity

- *Alpha diversity, α (1),* richness and evenness within a site/sample
    - Typically choose 2-3 of 7 metrics, **visualized independently**
- **Beta diversity, β (1),** compositional differences between sites/samples
    - Typically *10s-1000s of components* (dissimilarity of taxa abundance), **visualized as 2 components of 1 or 2 ordination method**
- Gamma diversity, γ (1), total species diversity in a landscape, `γ = α × β`
- (Zeta diversity, ζ (2), the degree of overlap in the type of taxa present between a set of observed sites/samples, overly sensitive to selection and order of sites/samples)

.footnote[
(1) R. H. Whittaker (1960) Vegetation of the Siskiyou Mountains, Oregon and California. Ecological Monographs, 30, 279–338. doi:10.2307/1943563
<br>(2) Hui, C., & McGeoch, M. A. (2014). Zeta diversity as a concept and metric that unifies incidence-based biodiversity patterns. The American Naturalist, 184(5), 684-694.
]

---

# Common Beta Diversity Metrics

## Bray-Curtis Dissimilarity

$$d_{BC}(X,Y) = \frac{\sum_i |X_i - Y_i|}{\sum_i (X_i + Y_i)}$$

- Range: 0 (identical) to 1 (completely different)
- **Abundance-weighted**: high-abundance taxa matter more
- **Ignores zeros**: doesn't penalize shared absences
- **Compositional**: abundance ratios matter

---

# Method 1: PCoA (Technical Details)

## The Math

1. Start with distance matrix $\mathbf{D}$
2. Convert to **centered Gram matrix** $\mathbf{G}$
3. Eigenvalue decomposition: $\mathbf{G} = \mathbf{U}\boldsymbol{\Lambda}\mathbf{U}^T$
4. **Principal coordinates** = $\mathbf{U}\sqrt{\boldsymbol{\Lambda}}$

## Key Properties

✓ Works with **any distance metric** (Bray-Curtis, Aitchison, etc.)  
✓ Preserves **pairwise distances** (in lower dims)  
✓ **Unconstrained**: axes are not explained by metadata  
✗ First axis not guaranteed to explain most variance  
✗ Can have **negative eigenvalues** (metric-dependent)

---

# Method 2: Canonical Analysis of Principal Coordinates (CAP)

## What it does

- Takes a **distance matrix** + **metadata variables**
- Fits **constrained ordination**
- Axes are linear combinations of metadata
- Each axis explains conditional % variance

---

# Method 2: CAP (Technical Details)

## The Process

1. Start with distance matrix $\mathbf{D}$
2. Compute **PCoA** (axes from $\mathbf{D}$)
3. **Constrain** PCoA axes to metadata via **RDA**
4. Retain axes that maximize correlation with constraints

## Key Properties

✓ **Interpretable axes**: tied to metadata  
✓ Tests whether metadata explains composition differences  
✓ Can use **PERMANOVA** to test significance  
✗ Requires metadata to be meaningful  
✗ Axes may explain less variance than unconstrained methods  

---

# Jie et al. 2017: Analysis Pipeline

```
1. Load abundance data (MetaPhlAn species profiles)
   ↓
2. Calculate Bray-Curtis distance matrix
   ↓
3. PCoA ordination (unconstrained)
   ↓
4. Visualize: PC1 vs PC2 colored by ACVD status
   ↓
5. PERMANOVA: Test if ACVD explains composition
   ↓
6. Beta-dispersion: Test if groups have different spread
```

---

# Jie et al. 2017: Results Interpretation

## PCoA Plot (Bray-Curtis)

```{r, eval=FALSE, echo=TRUE}
# From ACVD_reprex.Rmd (line 515)
physeq.beta <- phyloseq::ordinate(
  full_meta, 
  method = 'PCoA', 
  distance = 'bray'
)

pca.plot <- ggplot(pca.df) +
  geom_point(aes(x = PCoA1, y = PCoA2, color = ACVD), size = 2.5) +
  stat_ellipse(aes(x = PCoA1, y = PCoA2, color = ACVD), 
               level = 0.99) +
  xlab(paste0("PC1 (", round(percent_explained[1], 1), "%)"))
```

---

# Jie et al. 2017: Beta-Dispersion

## Are groups equally "spread out"?

```{r, eval=FALSE, echo=TRUE}
# From ACVD_reprex.Rmd (line 575)
beta_dispersion1 <- vegan::betadisper(
  dist_matrix, 
  beta_metadata$ACVD
)

anova_betadisper1 <- vegan::permutest(
  beta_dispersion1, 
  pairwise = FALSE
)
```

**Tests**: Do healthy vs ACVD groups have equal within-group distances?

- **Significant result**: Unequal dispersion
- **Interpretation**: Healthy samples more/less variable in composition

---

# PCA vs PCoA vs CAP: Summary

| Task | Best Method | Reason |
|------|-------------|--------|
| Quick exploration | **PCA** | Fast, intuitive, directly interpretable |
| Compare with Bray-Curtis | **PCoA** | Works with ecologically meaningful distance |
| Test if metadata explains beta diversity | **CAP** | Hypothesis-driven, PERMANOVA-testable |
| Compositional robustness | **PCoA + Aitchison** | Handles sum-to-one constraint |
| Publication-quality beta diversity plots | **PCoA** | Standard in microbiome literature |

