---
title: "Beta Diversity & Ordination Spaces"
subtitle: "From Distance Matrices to PCoA and CAP"
author: "Microbiome Analysis Workshop"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, default-fonts]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
editor_options: 
  chunk_output_type: console
---

```{r, eval=TRUE, echo=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, ## Code
  include = TRUE,
  message = FALSE, warning = FALSE, error = FALSE,
  out.height = "70%", fig.align = "center", fig.dpi = 300,
  cache = FALSE, cache.lazy = FALSE
)
```

```{r}
## Packages
suppressPackageStartupMessages({suppressWarnings({
  library(dplyr)
  library(ggplot2)
  # library(phyloseq)
  # library(vegan)
})})
options(tidyverse.quiet = TRUE)
theme_set(theme_bw())
here::i_am("BetaSpaces_Presentation.Rmd")

## Load artifacts
.verb <- ifelse(interactive(), TRUE, FALSE)
load(file = here::here(
  "./zScratchpad/ACVD_Inputs/outputs/ns_diversity_spaces.rda"), verbose = .verb)
```

# Terminology

- Mixed audiences; write Biology and interpret to Statistics

```{r, echo=FALSE}
tribble(
  ~Term,             ~Alias,                     ~Notaion_dimension,
  "Sample",          "Site, Observation, Row",   "n",
  "Variable",        "Colunm, feature, Metric",  "p", # "Dimension",
  "Abundance table", "Data",                     "n x p",
  "Meta data",       "Demographic data",         "n rows",
  "Distance",        "Dissimilarity",            "n x ~n*",
  "Ordination",      "Embedding, reduced scape", "n x d, where d <= ~n"
) %>%
  knitr::kable() %>%
  kableExtra::footnote(
    general = "* Biological distances sometimes don't meet the triangle inequality, leading to negative some eigenvalues. By default, most implementations drop negative eigenvalued components rather than diluting the others by adding the difference to all eigenvalues.",
    #general_title = "", 
    fixed_small_size = TRUE)
```

---

# Ecological Taxonomic Diversity

- *Alpha diversity(1), α,* richness and evenness within a site/sample
    - Typically choose 2-3 of 7 metrics, **visualized independently**
- **Beta diversity(1), β,** compositional differences between sites/samples
    - Typically *10s-1000s of components* (dissimilarity of taxa abundance), **visualized as 2 components of 1 or 2 ordination method**

.footnote[
(1) R. H. Whittaker (1960) Vegetation of the Siskiyou Mountains, Oregon and California. Ecological Monographs, 30, 279–338. doi:10.2307/1943563
<br><br><br>
]

---

# Ecological Taxonomic Diversity

- *Alpha diversity(1), α,* richness and evenness within a site/sample
    - Typically choose 2-3 of 7 metrics, **visualized independently**
- **Beta diversity(1), β,** compositional differences between sites/samples
    - Typically *10s-1000s of components* (dissimilarity of taxa abundance), **visualized as 2 components of 1 or 2 ordination method**
- Gamma diversity(1), γ, total species diversity in a landscape, `γ = α × β`
- (Zeta diversity(2), ζ, the degree of overlap in the type of taxa present between a set of observed sites/samples, overly sensitive to selection and order of sites/samples)

.footnote[
(1) R. H. Whittaker (1960) Vegetation of the Siskiyou Mountains, Oregon and California. Ecological Monographs, 30, 279–338. doi:10.2307/1943563

(2) Hui, C., & McGeoch, M. A. (2014). Zeta diversity as a concept and metric that unifies incidence-based biodiversity patterns. The American Naturalist, 184(5), 684-694.
]

---

# Alpha diversity, α

- Richness describes the unique number of taxa
- Evenness describes the distribution of taxa abundances
- Diversity is a measure of both richness and evenness

```{r, echo=FALSE}
tribble(
  ~Metric,      ~Class,      ~Context,  ~Description,
  "Observed",   "Richness",  "Wide",    "Count of observed taxa. Sensitive to rare taxa",
  "Shannon",    "Diversity", "Wide",    "Eveness and richness, log-based, range [0, log(n_tax)]",
  "Simpson",    "Evenness",  "Wide",    "Prob two randomly selected individuals are different species, range [0, 1]",
  "InvSimpson", "Evenness",  "+1",      "Inverse of Simpson (1/D). More intuitive scale",
  "Chao1",      "Richness",  "Non-ASV", "Accounting for unseen rare species, better for undersampled data",
  "ACE",        "Richness",  "Non-ASV", "Abundance-based Chao estimator, weights by abundance",
  "Fisher",     "Diversity", "?",       "Parametric index balancing richness and evenness, assumes log-series") %>% knitr::kable()
```

---

# The data

Follow the gut microbiome analysis in Jie et al, 2017 (3). Observed 385 samples of the gut microbiome in individuals with Artherosclerotic CardioVascular Disease (ACVD, "heart disease") compared to healthy individuals in a Chinese cohort. For our purpose, **goal is to visualize for taxa associated differntially by heart disease.**

- Data: 370 obs x 2397 taxa (relative abundances) -- `otu_table(physeq)`
- Metadata: 370 obs x 9 variables (Heart Disease, Sex, Age, Triglycerides, Cholesterols) -- `sample_data(physeq)`
- Alpha diversity: 370 obs x 4 within-site measures of richness, evenness, and diversity -- `estimate_richness(physeq, measures = c("Observed", "Shannon", "InvSimpson", "Fisher")`


.footnote[
(3) Jie, Z., Xia, H., Zhong, S. L., Feng, Q., Li, S., Liang, S., ... & Kristiansen, K. (2017). The gut microbiome in atherosclerotic cardiovascular disease. Nature communications, 8(1), 845.
]

---

# Alpha space

.pull-left[
Typically 2 to 7 alpha diversity metrics, visualized like...

````
estimate_richness(...) %>%
  as_tibble(rownames = "SampleID") %>%
  left_join(metadata) %>%
  pivot_longer() %>%
  ggplot() + ...
````

Apply apropriate  statistial test, *eg.* T, Wilcoxon, Kurskal-Wallis, or Baysian variants
]

.pull-right[
```{r, echo=F, fig.dpi=300}
# viridis::viridis(2, option = "A", begin = 0.3, end = 0.7) %>% dput()
acvd_colors <- c("#641A80FF", "#F76F5CFF")
ggplot(jia_alpha_long) +
  geom_boxplot(aes(x = ACVD, y = value, fill = ACVD)) +
  ylab('Alpha Diversity') +
  scale_fill_manual(values = acvd_colors) +
  facet_wrap(~Diversity_Measure, scales = 'free') +
  theme_bw()
```
]


---

# Alpha space 2

.pull-left[
## Recomendations

1. Move from `geom_boxplot()` to `geom_violin() + geom_jitter()`
1. `viridis` palette was relatively color-blind friendly, alternatively color-blind 
friendly palettes with `scale_color_brewer()` and `scale_fill_brewer()`
1. Consider adding arrows to indicate domain prefered directions, esp when 
audiences may not be familar with all metrics
]

.pull-right[
```{r}
y_ranges <- jia_alpha_long |>
  group_by(Diversity_Measure) |>
  summarise(y_min = min(value, na.rm = TRUE),
            y_max = max(value, na.rm = TRUE),
            .groups = "drop") |>
  mutate(
    arrow_y_start = y_min + 0.05 * (y_max - y_min),
    arrow_y_end = y_max - 0.05 * (y_max - y_min))

ggplot(jia_alpha_long) +
  geom_violin(aes(x = ACVD, y = value, fill = ACVD), alpha = .3) +
  geom_jitter(aes(x = ACVD, y = value, color = ACVD)) +
  ## Arrows
  geom_segment(
    data = y_ranges,
    aes(x = .5, xend = .5, y = arrow_y_start, yend = arrow_y_end),
    arrow = arrow(ends = "last", length = unit(0.3, "cm"), type = "closed"),
    color = "green4", size = 1, inherit.aes = FALSE) +
  ylab('Alpha Diversity') +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  facet_wrap(~Diversity_Measure, scales = 'free') +
  theme_bw()
```
]

---

# Alpha space 3

.pull-left[
## Recomendations

- More information when pairing metrics:
    - **Visualize where evenness and richness disagree**
    - 2d Scatter plot with tooltips on sample/site & metadata with `plotly::ggplotly()`
    - Scatterplot matrix(4) (aka SPLOM, pairs plot), `GGally::ggpairs()` (right)


.footnote[
(4) Hartigan, J. A. (1975). Clustering algorithms.
John Wiley & Sons, Inc..
]
]

.pull-right[
```{r}
GGally::ggpairs(jia_alpha_wide, columns = 2:5,
                aes(color = ACVD, fill = ACVD, shape = ACVD)) +
  scale_color_brewer(palette = "Dark2")
```
]


---

# Beta diversity

TODO

Distance (`vegdist(otu_table(physeq), method = "bray")`): Bray-Curtis dissimilarity, 2395 x 2296 taxa
Ordination PCoa (`ordinate(physeq, distance = "bray", method = "PCoA")`): 370 obs x 211 PCoA Components


<!-- with a particular focus on Akkermansia sp. -->

-----

.footnote[
Jie, Z., Xia, H., Zhong, S. L., Feng, Q., Li, S., Liang, S., ... & Kristiansen, K. (2017). The gut microbiome in atherosclerotic cardiovascular disease. Nature communications, 8(1), 845. 
(1805 citations 6/26/2026) 
]

---

# Common Beta Diversity Metrics

## Bray-Curtis Dissimilarity

$$d_{BC}(X,Y) = \frac{\sum_i |X_i - Y_i|}{\sum_i (X_i + Y_i)}$$

- Range: 0 (identical) to 1 (completely different)
- **Abundance-weighted**: high-abundance taxa matter more
- **Ignores zeros**: doesn't penalize shared absences
- **Compositional**: abundance ratios matter

---

# Common Beta Diversity Metrics (cont.)

## Aitchison Distance

$$d_{A}(X,Y) = \sqrt{\sum_i \left(\log\frac{X_i}{g(X)} - \log\frac{Y_i}{g(Y)}\right)^2}$$

Where $g(X) = \text{geometric mean}$ of $X$

- **CLR-transformed**: centered log-ratio
- Accounts for **compositional nature** of microbiome data
- Better for **sparse, high-dimensional** data
- Often more powerful in differential abundance tests

---

# The Distance Matrix

## What is it?

A **symmetric matrix** of pairwise distances between all samples.

```{r, eval=FALSE, echo=TRUE}
# Example with 6 samples
dist_matrix <- 
  ┌─────┬──────┬──────┬──────┬──────┐
  │ 0   │ 0.32 │ 0.45 │ 0.28 │ 0.61 │
  ├─────┼──────┼──────┼──────┼──────┤
  │ 0.32│ 0    │ 0.38 │ 0.29 │ 0.55 │
  ├─────┼──────┼──────┼──────┼──────┤
  │ 0.45│ 0.38 │ 0    │ 0.52 │ 0.18 │
  ├─────┼──────┼──────┼──────┼──────┤
  │ 0.28│ 0.29 │ 0.52 │ 0    │ 0.64 │
  ├─────┼──────┼──────┼──────┼──────┤
  │ 0.61│ 0.55 │ 0.18 │ 0.64 │ 0    │
  └─────┴──────┴──────┴──────┴──────┘
```

**Problem**: High-dimensional! Can't visualize 6 samples in their true space.

**Solution**: Ordination → 2D representation preserving these distances.

---

# Ordination: Goal

## Map high-dimensional distances to 2D (or 3D)

```{r setup, include=FALSE}
library(ggplot2)
library(vegan)
library(tidyverse)

set.seed(2847)
# Create synthetic distance example
dist_example <- as.matrix(dist(matrix(rnorm(18), ncol=3)))
```

**Original distances** → **2D coordinates**

- Minimize information loss (-- Really?)
- Preserve relative distances
- Interpretable axes (-- Really?)

---

# Method 1: Principal Coordinates Analysis (PCoA)

## What it does

- Takes a **distance matrix** as input
- Decomposes it via eigenvalue decomposition
- Returns **new axes** (principal coordinates)
- Each axis explains % of total variance

---

# Method 1: PCoA (Technical Details)

## The Math

1. Start with distance matrix $\mathbf{D}$
2. Convert to **centered Gram matrix** $\mathbf{G}$
3. Eigenvalue decomposition: $\mathbf{G} = \mathbf{U}\boldsymbol{\Lambda}\mathbf{U}^T$
4. **Principal coordinates** = $\mathbf{U}\sqrt{\boldsymbol{\Lambda}}$

## Key Properties

✓ Works with **any distance metric** (Bray-Curtis, Aitchison, etc.)  
✓ Preserves **pairwise distances** (in lower dims)  
✓ **Unconstrained**: axes are not explained by metadata  
✗ First axis not guaranteed to explain most variance  
✗ Can have **negative eigenvalues** (metric-dependent)

---

# Method 2: Canonical Analysis of Principal Coordinates (CAP)

## What it does

- Takes a **distance matrix** + **metadata variables**
- Fits **constrained ordination**
- Axes are linear combinations of metadata
- Each axis explains conditional % variance

---

# Method 2: CAP (Technical Details)

## The Process

1. Start with distance matrix $\mathbf{D}$
2. Compute **PCoA** (axes from $\mathbf{D}$)
3. **Constrain** PCoA axes to metadata via **RDA**
4. Retain axes that maximize correlation with constraints

## Key Properties

✓ **Interpretable axes**: tied to metadata  
✓ Tests whether metadata explains composition differences  
✓ Can use **PERMANOVA** to test significance  
✗ Requires metadata to be meaningful  
✗ Axes may explain less variance than unconstrained methods  

---

# Method 3: Principal Component Analysis (PCA)

## What it does

- Takes **abundance table** directly
- No distance metric required
- Finds **axes of maximum variance** in abundance space
- Common for exploratory analysis

---

# Comparison: nxp Data Processing

|   | PCA | LDA | PCoA | CAP |
|---|-----|-----|------|-----|
| **Input** | Abundance table (n × p) | Abundance table + metadata | Distance matrix (n × n) | Distance matrix + metadata |
| **Distance metric** | None (Euclidean implicit) | None (Euclidean implicit) | Any (Bray, Aitchison, etc.) | Any (Bray, Aitchison, etc.) |
| **Eigenvalues** | Always ≥ 0 | Always ≥ 0 | Can be negative | Can be negative |
| **First axis captures** | Max variance in abundance | Separation between clusters | Best approximate distances | Metadata-constrained variance |
| **Handles compositional data** | Not well | Not well | Yes (metric-dependent) | Yes (metric-dependent) |
| **Requires metadata** | No |  Yes (required) | No | Yes (required) |
| **Linear basis** | Yes | Yes | No (projects sites somehow) | No (projects sites somehow) |

---

# PCA: The Old School

## n samples × p taxa

```
┌───────────────────────────────────┐
│ Sample₁: [0.3, 0.1, 0.5, ...]     │
│ Sample₂: [0.2, 0.4, 0.3, ...]     │
│ Sample₃: [0.1, 0.2, 0.6, ...]     │
│ ...                               │
└───────────────────────────────────┘

PCA:
┌────────────────────────────────────────────┐
│ Find axes maximizing variance in original  │
│ abundance dimensions                       │
│                                            │
│ PC1 = 0.5 × taxa₁ + 0.3 × taxa₂ + ...    │
│ PC2 = -0.2 × taxa₁ + 0.6 × taxa₃ + ...   │
└────────────────────────────────────────────┘
```

**Problem**: Ignores **compositional nature** of microbiome (sum-to-one constraints)

---

# PCoA: From Distances

## Distance matrix (n × n)

```
      S1    S2    S3
S1  [ 0.0  0.32  0.45 ]
S2  [ 0.32  0.0  0.38 ]
S3  [ 0.45  0.38  0.0 ]

PCoA:
┌──────────────────────────────────────────────┐
│ Decompose distance matrix via eigenanalysis  │
│ Preserve pairwise distances in lower dims    │
│                                              │
│ PC1: [-0.25, 0.10, 0.15]  (explains 45%)    │
│ PC2: [0.18, -0.32, 0.14]  (explains 30%)    │
└──────────────────────────────────────────────┘
```

**Advantage**: Works with **any distance metric**, including compositionally-aware ones

---

# CAP: Adding Metadata

## Distance matrix + constraints

```
      S1    S2    S3
S1  [ 0.0  0.32  0.45 ]
S2  [ 0.32  0.0  0.38 ]
S3  [ 0.45  0.38  0.0 ]

Metadata:
S1: ACVD=0, Age=45
S2: ACVD=1, Age=52
S3: ACVD=0, Age=48

CAP:
┌─────────────────────────────────────────────┐
│ Project samples to metadata-constrained     │
│ ordination space using RDA                  │
│                                             │
│ CAP1: Explains ACVD + Age (67% of variance)│
│ CAP2: Residual variation (20%)              │
└─────────────────────────────────────────────┘
```

**Advantage**: Test whether metadata predicts composition

---

# Jie et al. 2017: Current Practice

## Study Design

- **Samples**: 385 Chinese individuals
- **Groups**: Healthy vs ACVD (artherosclerotic CVD)
- **Sequencing**: MetaPhlAn species-level abundances
- **Beta diversity metric**: Bray-Curtis

---

# Jie et al. 2017: Analysis Pipeline

```
1. Load abundance data (MetaPhlAn species profiles)
   ↓
2. Calculate Bray-Curtis distance matrix
   ↓
3. PCoA ordination (unconstrained)
   ↓
4. Visualize: PC1 vs PC2 colored by ACVD status
   ↓
5. PERMANOVA: Test if ACVD explains composition
   ↓
6. Beta-dispersion: Test if groups have different spread
```

---

# Implementation: phyloseq

## Workflow in R



---

# Implementation: CAP in phyloseq

## Constrained Ordination

```{r, eval=FALSE, echo=FALSE}
# CAP with metadata constraints
cap_result <- phyloseq::ordinate(
  physeq,
  method = "CAP",
  distance = "bray",
  formula = ~ ACVD + Age + Gender
)

# Extract variance explained
# (shows % explained by constraints)
cap_result$inertia$constrained_perc

# Test significance (PERMANOVA)
anova(cap_result)

# Plot
plot_ordination(physeq, cap_result, color = "ACVD")
```

---

# Implementation: Aitchison Distance

## Compositionally-aware PCoA

```{r, eval=FALSE, echo=FALSE}
# Using microViz for modern workflow
physeq |>
  tax_transform("identity", rank = "Species") |>
  dist_calc(dist = "aitchison") |>
  ord_calc("PCoA") |>
  ord_plot(color = "ACVD")

# Under the hood:
# 1. CLR-transform abundances
# 2. Compute Euclidean distance
# 3. PCoA on Euclidean distances
```

---

# Jie et al. 2017: Results Interpretation

## PCoA Plot (Bray-Curtis)

```{r, eval=FALSE, echo=FALSE}
# From ACVD_reprex.Rmd (line 515)
physeq.beta <- phyloseq::ordinate(
  full_meta, 
  method = 'PCoA', 
  distance = 'bray'
)

pca.plot <- ggplot(pca.df) +
  geom_point(aes(x = PCoA1, y = PCoA2, color = ACVD), size = 2.5) +
  stat_ellipse(aes(x = PCoA1, y = PCoA2, color = ACVD), 
               level = 0.99) +
  xlab(paste0("PC1 (", round(percent_explained[1], 1), "%)"))
```

**Finding**: Modest separation by ACVD status, ellipses overlap considerably

---

# Jie et al. 2017: Statistical Testing

## PERMANOVA Results

```{r, eval=FALSE, echo=TRUE}
# From ACVD_reprex.Rmd (line 570)
permanova_result1 <- vegan::adonis2(
  dist_matrix ~ ACVD + Age + Gender,
  method = "bray", 
  data = beta_metadata
)
```

**Findings**:
- ACVD significant (p < 0.05) 
- **R² = 2%**: Only 2% of variance explained
- Age also significant
- Gender: not significant
- **Interpretation**: Differences exist but are small

---

# Jie et al. 2017: Beta-Dispersion

## Are groups equally "spread out"?

```{r, eval=FALSE, echo=TRUE}
# From ACVD_reprex.Rmd (line 575)
beta_dispersion1 <- vegan::betadisper(
  dist_matrix, 
  beta_metadata$ACVD
)

anova_betadisper1 <- vegan::permutest(
  beta_dispersion1, 
  pairwise = FALSE
)
```

**Tests**: Do healthy vs ACVD groups have equal within-group distances?

- **Significant result**: Unequal dispersion
- **Interpretation**: Healthy samples more/less variable in composition

---

# When to Use Each Method

## Quick Decision Tree

```
                ┌─ Input: Abundance table
                │         └─ PCA (exploratory)
        Start ──┤
                └─ Input: Distance matrix
                   ├─ No metadata → PCoA (unconstrained)
                   └─ Have metadata
                      └─ Test metadata effect? 
                         ├─ Yes → CAP (hypothesis test)
                         └─ Visualize only → PCoA
```

---

# PCA vs PCoA vs CAP: Summary

| Task | Best Method | Reason |
|------|-------------|--------|
| Quick exploration | **PCA** | Fast, intuitive, directly interpretable |
| Compare with Bray-Curtis | **PCoA** | Works with ecologically meaningful distance |
| Test if metadata explains beta diversity | **CAP** | Hypothesis-driven, PERMANOVA-testable |
| Compositional robustness | **PCoA + Aitchison** | Handles sum-to-one constraint |
| Publication-quality beta diversity plots | **PCoA** | Standard in microbiome literature |

---

# Key Takeaways

### Beta Diversity Measures Composition Differences

- Bray-Curtis: Traditional, abundance-weighted
- Aitchison: Compositionally-aware, recommended for sparse data

### Ordination Maps High-D Distances to 2D

- **PCA**: Works on abundance table directly
- **PCoA**: Works on any distance metric
- **CAP**: Adds metadata constraints for hypothesis testing

### Jie et al. 2017 Pattern

- Used **Bray-Curtis + PCoA** (current standard)
- Small R² in PERMANOVA (2% variance explained)
- Significant but modest compositional differences between groups

---

# Resources & References

### Key Papers

- **Jie et al. 2017**: doi: 10.1038/s41467-017-00900-1
- **Aitchison (1986)**: Statistical Analysis of Compositional Data
- **Ramette (2007)**: Multivariate analyses in microbial ecology

### R Packages

- `phyloseq`: Microbiome data structures & plotting
- `vegan`: Ordination methods (PCoA, CAP, PERMANOVA)
- `microViz`: Modern tidy workflows

### This Workshop

- **ACVD_reprex.Rmd**: Full reproducible example
- **1a_vegan_ordination.R**: Utility functions

---
class: center, middle

# Questions?

---

# Appendix: Behind the PCoA Curtain

## Eigenvalue Interpretation

```{r, eval=FALSE, echo=TRUE}
# PCoA eigenvalues show variance captured
eigenvalues <- pcoa_result$values$Eigenvalues
percent_explained <- eigenvalues / sum(eigenvalues) * 100

# Example output:
# PC1: 8.2% of variance explained
# PC2: 5.1% of variance explained
# PC3: 4.3% of variance explained
# ...

# Note: Not necessarily decreasing!
# Because we're preserving pairwise distances,
# not maximizing variance like PCA does.
```

---

# Appendix: CAP vs RDA

## Relationship

**CAP** = **RDA** applied to **PCoA axes**

```
Distance Matrix
       ↓
    PCoA ←─── Eigenvalue decomposition
       ↓
   PCoA axes (unconstrained)
       ↓
    RDA ←─── Constrain to metadata
       ↓
  CAP axes (constrained)
```

- RDA: Directly on abundance-derived distance
- CAP: On PCoA axes of user-specified distance

**Result**: CAP more flexible (any distance metric)
