---
title: "Metabolic health in Chinese cohorts: gut microbiome of cardiovascular disease"
#subtitle:
author: "NS slimmed from Madison Youngblom (madison.youngblom@iff.com)"
date: "`r format(Sys.time(), '%b %d, %Y')`"
output:
  html_document:
    #includes:
    #  before_body: assets/header-footer.html
    #css: assets/styles.css
    toc: true
    toc_depth: 3
    toc_float: true
    self_contained: yes
    code_folding: hide
    code_tools: true
editor_options:
  chunk_output_type: console
---

# 00 Setup

## 00.1 Inputs

```{r inputs, include=FALSE}
{
  tictoc::tic("Compile whole report")
  project <- 'NA'
  proj_desc <- 'Re-analysis of Jie et al, 2017 (10.1038/s41467-017-00900-1) data looking at the gut microbiome in individuals with artherosclerotic cardiovascular disease (ACVD) compared to healthy individuals in a Chinese cohort. Looking for taxa associated with metabolic health with a particular focus on Akkermansia sp.'
  nsamples <- '385'
  seq_loc <- 'NA'
  seqread <- 'PRJEB21528'
  
  ## Input filenames
  proj_path <- "./zScratchpad/ACVD_Inputs/"
  metadata_file <- paste0(proj_path, 'inputs/Jieetal_2017_supp.txt')
  read_summary <- paste0(proj_path, 'inputs/Read_counts_quality.txt')
  metaphlan_unclass <-
    paste0(proj_path, 'inputs/merged_metaphlan_count_species_wUnclass.txt')
  stopifnot(file.exists(c(metadata_file, read_summary, metaphlan_unclass)))
}
```

## 00.2 Options and settings

```{r setup, include=FALSE}
## Set higher level report parameters
{
  DO_CACHE <- FALSE ## Cache chunks? If you see stale content change to FALSE
  .context <- paste0("Compiled: ", Sys.Date(), 
                     ", with report parameteres: ",
                     "DO_CACHE = ", DO_CACHE#, ", "
                     ## Add new parameters here
  )
}

## Options
options(tidyverse.quiet = TRUE)
#theme_set(theme_bw())
chunk_time <- dplyr::tibble(chunk = NULL, seconds = NULL)
#knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(
  echo = TRUE, ## Code
  include = TRUE,
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  out.width = "100%",
  fig.align = "left",
  # out.height = "500px",
  # fig.width=ggplot2::unit(6.5, "in"),
  # fig.height=ggplot2::unit(6.5/1.61, "in"),
  # fig.width=ggplot2::unit(6.5, "in"),
  # fig.height=ggplot2::unit(6.5/1.61, "in"),
  # out.height = "500px",
  # out.width = "700px"
  cache = DO_CACHE, cache.lazy = DO_CACHE
)

## Report function util
# proj_dir <- function(path) {
#   if (interactive()) {
#     return(path)
#   } else {
#     return(paste0("../", path))
#   }
# }
# my_kbl <- function(
    #     tbl, digits = getOption("digits"), align = NULL, font_size = NULL, ...) {
#   tbl %>%
#     knitr::kable(
#       booktabs = TRUE, digits = digits, align = align
#     ) %>%
#     kableExtra::kable_styling(
#       bootstrap_options = "striped", latex_options = "striped",
#       full_width = FALSE, font_size = font_size, ...
#     )
# }
```

## 00.3 Packages

```{r packages, include=FALSE}
if(!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if(!require("maaslin3", quietly = TRUE))
  BiocManager::install("maaslin3")
if(!require("microViz", quietly = TRUE))
  pak::pak("david-barnett/microViz")

suppressPackageStartupMessages({suppressWarnings({
  library(microbiome)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(RColorBrewer)
  library(stringr)
  library(ggpubr)
  library(DT)
  library(knitr)
  library(kableExtra)
  library(plotly)
  library(grid)
  library(shiny)
  library(viridis)
  library(maaslin3)
  library(tibble)
  library(microViz)
  library(phyloseq)
})})

my_gg_theme <- theme_bw() +
  theme(axis.text = element_text(size = 10),
        axis.title = element_text(size = 10,face = 'bold'),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10, face = 'bold'))
acvd_colors <- viridis(2, option = "A", begin = 0.3, end = 0.7)
```

## 00.4 Functions

```{r functions, include=FALSE}
metaphlan_physeq <- function(
    metadata_file, metaphlan_file, sample_header = 'SRA', rm_samples = c()
){
  ##Import metadata
  df_metadata <- read.delim(
    metadata_file, header = TRUE, sep = "\t", fill = TRUE,
    na.strings = c("", "NA")) %>%
    rename('SampleID' = {{sample_header}},
           "TRIG" = "TRIG..mmol.L.",
           "LDLC" = "LDLC..mmol.L.",
           "CHOL" = "CHOL..mmol.L.",
           "HDLC" = "HDLC..mmol.L.",
           "ACVD" = "ACVD.status",
           "Age" = "Age..year.",
           "BMI" = "Body.Mass.Index..BMI.") %>%
    select(SampleID, Gender, Age, ACVD, TRIG, LDLC, HDLC, CHOL, BMI) %>%
    filter(!SampleID %in% rm_samples)
  #df_metadata <- df_metadata[sample(nrow(df_metadata), 50),]
  
  rownames(df_metadata) <- df_metadata$SampleID
  META <- phyloseq::sample_data(df_metadata)
  
  ##Import metaphlan
  df_metaphlan <- read.delim(
    metaphlan_file, header = TRUE, row.names = 1, sep = "\t",
    check.names = FALSE, skip = 1) %>%
    rename_with(~ str_remove_all(., ".metaphlan"), .cols = everything()) %>%
    select(!all_of(rm_samples))
  
  otu <- data.matrix(df_metaphlan)
  OTU <- phyloseq::otu_table(otu, taxa_are_rows = TRUE)
  
  rownames <- rownames(df_metaphlan)
  #Split rownames by | into a new data frame:
  taxmat <- data.frame(stringr::str_split_fixed(rownames, "\\|",7))
  colnames(taxmat) = c(
    "Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
  rownames(taxmat) <- rownames(df_metaphlan)
  taxmat <- apply(taxmat, c(1, 2), function(x) gsub("^.*__", "", x))
  taxmat <- as.matrix(taxmat)
  TAX = phyloseq::tax_table(taxmat)
  physeq = phyloseq::phyloseq(META, OTU, TAX)
  return(physeq)
}


calculate_da_genOdds <- function(
    df, meta, treat = "Treatment", assume_no_effect = FALSE
){
  df.res <- data.frame(
    bin = vector(length = ncol(df)),
    estimate = vector(length = ncol(df)),
    std.error = vector(length = ncol(df)),
    p.value = vector(length = ncol(df))
  )
  
  meta <- meta[order(meta$Sample_ID), ]
  df <- df[order(rownames(df)), , drop = F]
  
  for (s in 1:ncol(df)) {
    taxa <- names(df)[s]
    if (all(meta$Sample_ID != rownames(df))) {
      stop("Metadata and Spec abund table are not in the same order")
    } else {
      spec.ab <- df[, s]
      tmp <- cbind.data.frame(spec.ab, meta) # build df with species and metadata
      x <- try(lo <- genodds(
        tmp$spec.ab, tmp[, treat], assume_no_effect = assume_no_effect
      )) # in case of error
      if (class(x)[1] == "try-error") {
        df.res[s, "bin"] <- spec
        df.res[s, "estimate"] <- NA
        df.res[s, "std.error"] <- NA
        df.res[s, "p.value"] <- NA
      } else {
        df.res[s, "bin"] <- spec
        df.res[s, "estimate"] <- lo$pooled_lnodds
        df.res[s, "std.error"] <- lo$pooled_SElnodds
        df.res[s, "p.value"] <- lo$pooled_p
      }
    }
  }
  return(df.res)
}

plot_heatmap_may <- function (
    x, subset.top, transformation, VariableA, heatcolors = NULL, ...
){
  topOTU <- phyobj1 <- phyobj2 <- otu.mat <- NULL
  meta.tab <- select.meta <- color.heatmap <- NULL
  x <- suppressWarnings(suppressMessages(microbiomeutilities::format_to_besthit(x)))
  if (!is.null(subset.top)) {
    message(paste0("Top ", subset.top, " OTUs selected", 
                   sep = " "))
    topOTU <- top_taxa(x, n = subset.top)
  }
  else {
    stop("specify a number/value for subset.top")
  }
  if (transformation == "log10") {
    message("log10, if zeros in data then log10(1+x) will be used")
    phyobj1 <- prune_taxa(topOTU, x)
    message("First top taxa were selected and \nthen abundances tranformed to log10(1+X)")
    phyobj2 <- transform(phyobj1, "log10")
  }
  else if (transformation == "compositional") {
    phyobjx <- transform(x, "compositional")
    phyobj2 <- prune_taxa(topOTU, phyobjx)
    message("First converted to compositional \n then top taxa were selected")
  }
  else if (transformation == "Z-OTU") {
    phyobj1 <- prune_taxa(topOTU, x)
    phyobj2 <- transform(phyobj1, "Z")
    message("First top taxa were selected and \nthen abundances tranformed to Z values")
  }
  else if (transformation == "clr") {
    phyobj1 <- prune_taxa(topOTU, x)
    phyobj2 <- transform(phyobj1, "clr")
    message("First top taxa were selected and \nthen abundances tranformed to clr")
  }
  else if (!is.null(transformation)) {
    stop("specify a number for transformation, log10, compositional, Z-OTU, clr")
  }
  otu.mat <- abundances(phyobj2)
  meta.tab <- meta(phyobj2)
  select.meta <- subset(meta.tab, select = c(VariableA))
  if (is.null(heatcolors)) {
    color.heatmap <- brewer.pal(6, "Spectral")
  }
  else {
    color.heatmap <- heatcolors
  }
  newnames0 <- gsub(x = str_split_fixed(rownames(otu.mat), 
                                        pattern = '\\:', n = 2)[,2],
                    pattern = 'g__',
                    replacement = '')
  newnames <- NULL
  newnames <- lapply(newnames0, function(x) bquote(italic(.(x))))
  heatmap <- pheatmap::pheatmap(otu.mat, labels_row = as.expression(newnames), 
                                annotation_col = select.meta, color = color.heatmap, ...)
  return(list(plot = heatmap, tax_tab = otu.mat))
}
```

# 0 Abstract -- (hard coded)

## 0.1 Methods

Re-analysis of Jie et al, 2017 (10.1038/s41467-017-00900-1) data looking at the gut microbiome in individuals with artherosclerotic cardiovascular disease (ACVD) compared to healthy individuals in a Chinese cohort. Looking for taxa associated with metabolic health with a particular focus on _Akkermansia_ sp.

## 0.2 Results

* Due to missing metadata, there was not enough power to detect significant associations between gut microbiome taxa and certain metadata (including obesity related markers such as BMI).
* Comparison of gut microbiome composition between healthy and [ACVD cohorts](#acvd-table) revealed a number of significant differences both at the species and genus levels.
* Akkermansia
* The prevalence of _A. muciniphila_ is [significantly higher](#akk-diff-table) in the ACVD cohort compared to healthy controls.
* The median abundance of _A. muciniphila_ and _A. massliensis_ are [lower](#akk) in the ACVD cohort compared to healthy controls (although this comparison is not significant).
* Note the low prevalence of [_A. guanxiensis_](#akk-box) in this cohort, which is a rare species of _Akkermansia_ that is associated with centenarians that has so far (as far as I can tell) only been isolated from Chinese individuals.


# 1 Project summary -- (hard coded)

```{r proj}
proj_df <- data.frame(
  Type = c('Project', 'Description', 'Number of Samples', 'Sequencing Facility',
           'SeqRead Project', 'Metadata', 'Data QC', 'Taxonomic Abundance'),
  Value = c(project, proj_desc, nsamples, seq_loc, seqread,
            basename(metadata_file), basename(read_summary),
            basename(metaphlan_unclass)))

kable(proj_df, col.names = NULL) %>%
  kable_styling() %>%
  column_spec(1, width = "2in")
```


# 2 Sequence Data {#seqdata}

<!-- ### 2.1 Read counts -- (read_summary) -->

<!-- ```{r seqdata} -->
<!-- seq <- read.delim(read_summary) %>% -->
<!--   mutate(RawReads = round(RawReads / 1e6,1), -->
<!--          TrimmedReads = round((TrimmedReads + TrimmedSingles) / 1e6,1), -->
<!--          FilteredReads = round((FilteredReads + FilteredSingles) / 1e6,1), -->
<!--          TrimRet = round((TrimmedReads/RawReads) * 100, 1), -->
<!--          FiltRet = round((FilteredReads/RawReads) * 100, 1) -->
<!--   ) %>% -->
<!--   select(Sample, -->
<!--          'Raw Reads (M)' = RawReads, -->
<!--          'Trimmed Reads (M)' = TrimmedReads, -->
<!--          'Retained after trimming (% raw)' = TrimRet, -->
<!--          'De-contaminated Reads (M)' = FilteredReads, -->
<!--          'Retained after filtering (% raw)' = FiltRet) -->

<!-- DT::datatable( -->
<!--   seq, rownames=FALSE, width='100%', -->
<!--   options = list( -->
<!--     scrollX = '500px', -->
<!--     scrollY = '300px', -->
<!--     paging = FALSE, -->
<!--     info = FALSE, -->
<!--     columnDefs = list(list( -->
<!--       targets = seq(1,5), # Column index to format -->
<!--       render = JS( -->
<!--         "function(data, type, row, meta) {", -->
<!--         "return type === 'display' ? data.toFixed(1) : data;", -->
<!--         "}" -->
<!--       ))) -->
<!--   )) %>% -->
<!--   formatStyle(columns = colnames(.$x$data), `font-size` = '14px') -->
<!-- ``` -->


<!-- ### 2.2 Quality metrics -- (read_summary) -->

<!-- ```{r fastqc, fig.cap="Figure 2.2: FastQC quality metrics."} -->
<!-- qc_colors <- c('green3','yellow2','orangered2') -->
<!-- fastqc <- read.delim(read_summary) %>% -->
<!--   select(all_of(c(1, seq(7, 16)))) -->
<!-- col.order <- colnames(fastqc) -->

<!-- fastqc.l <- fastqc %>% -->
<!--   pivot_longer(cols = -Sample, names_to = 'stat', values_to = 'value') %>% -->
<!--   mutate(stat = factor(stat, levels = col.order)) %>% -->
<!--   arrange(Sample) -->

<!-- fastqc_plot <- ggplot(fastqc.l) + -->
<!--   geom_tile(aes(x = stat, y = Sample, fill = value)) + -->
<!--   scale_fill_manual(values = qc_colors, breaks = c('PASS', 'WARN', 'FAIL')) + -->
<!--   scale_x_discrete(labels = gsub('\\.', ' ', col.order[-1])) + -->
<!--   scale_y_discrete(position = 'right') + -->
<!--   labs(x = "Sample", y = "Statistic") + -->
<!--   theme(axis.ticks.y = element_blank(), -->
<!--         axis.text.y = element_blank(), -->
<!--         #legend.position='none', -->
<!--         axis.title = element_blank(), -->
<!--         axis.text.x = element_text(angle = -25, hjust = 1)) -->

<!-- m <- list( -->
<!--   #l = 50, -->
<!--   #r = 50, -->
<!--   b = 50, -->
<!--   t = 50, -->
<!--   pad = 0.5) -->
<!-- div( -->
<!--   ggplotly(fastqc_plot, width = 800, height = 500) %>% -->
<!--     layout(autosize = F, margin = m), -->
<!--   align = 'left') -->
<!-- ``` -->

### 2.3 Reads mapping to unclassified taxa -- (metaphlan_unclass, metadata_file)

**This section creates `physeq` & `bad_samples`, consuming `metaphlan_unclass`, `metadata_file` (inputs/merged_metaphlan_count_species_wUnclass.txt, and inputs/Jieetal_2017_supp.txt respectively)**

A very high proportion of reads mapping to "unclassified" taxa could indicate a problem with the sample or the sequencing. Values between 10-30% are typical. Samples with >40% of reads mapping to unclassified taxa will be removed from further analysis.

```{r unclassified, fig.cap="Figure 2.3: Proportion of reads mapping to unclassified taxa by cohort. Points in red represent samples removed from analysis."}
df_unclass <- read.delim(metaphlan_unclass, header = TRUE, sep = "\t",
                         check.names = FALSE, skip = 1) %>%
  rename_with(~ str_remove_all(., ".metaphlan"), .cols = everything()) %>%
  column_to_rownames(var = 'clade_name') %>%
  mutate(across(everything(), ~ .x/ sum(.x))) %>%
  rownames_to_column(var = 'clade_name') %>%
  filter(clade_name == 'UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED') %>%
  column_to_rownames(var = 'clade_name')

metadata <- read.delim(metadata_file, header = TRUE, sep = "\t",
                       fill = TRUE, na.strings = c("","NA")) %>%
  rename(SampleID = "SRA",
         TRIG = "TRIG..mmol.L.",
         LDLC = "LDLC..mmol.L.",
         CHOL = "CHOL..mmol.L.",
         HDLC = "HDLC..mmol.L.",
         ACVD = "ACVD.status",
         Age = "Age..year.",
         BMI = "Body.Mass.Index..BMI.") %>%
  select(SampleID, Gender, Age, ACVD, TRIG, LDLC, HDLC, CHOL) %>%
  mutate(ACVD = factor(ACVD, levels = c("0", "1")))

unclass_wmeta <- as.data.frame(t(df_unclass)) %>%
  rownames_to_column(var = "SampleID") %>%
  left_join(metadata, by = "SampleID") %>%
  rename(UNCLASSIFIED = "UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED|UNCLASSIFIED") %>%
  mutate(over40 = case_when(UNCLASSIFIED >= 0.4 ~ "yes",
                            TRUE ~ "no"))

bad_samples <- unclass_wmeta %>%
  filter(UNCLASSIFIED >= 0.4) %>%
  pull(SampleID)

unclass.plot <- ggplot(unclass_wmeta, aes(x = ACVD, y = UNCLASSIFIED * 100)) +
  geom_boxplot(aes(fill = ACVD)) +
  geom_point(data = unclass_wmeta %>% filter(over40 == "yes"),
             aes(group = SampleID), color = "red") +
  scale_fill_manual(values = acvd_colors) +
  my_gg_theme +
  labs(y = "MetaPhlAn Unclassified (%)") +
  guides(color = "none")
unclass.plot
```

```{r metaphlan_physeq, include=FALSE}
physeq <- metaphlan_physeq(metadata_file = metadata_file,
                           metaphlan_file = metaphlan_unclass,
                           sample_header = 'SRA',
                           rm_samples = c(bad_samples, 'ERR2017627'))

# make factors
metadata <- data.frame(sample_data(physeq)) %>%
  mutate(ACVD = factor(ACVD, levels = c('0', '1')))
sample_data(physeq) <- metadata
#' @examples
#' print(physeq)
```


# 3 Analysis


## 3.1 Alpha diversity -- (physeq)

Alpha diversity measures the richness (number of taxa) and evenness (relative abundance) of samples. Observed diversity describes the richness while the Shannon index combines richness and evenness. There are no significant differences in alpha diversity between groups.

```{r alpha, warnings=FALSE, fig.cap="Figure 3.1: Alpha diversity (observed species count and Shannon index) plotted by ACVD status."}
full_meta <- microViz::ps_filter(physeq, !is.na(Gender) & !is.na(Age),
                                 .keep_all_taxa = TRUE)

alpha.df <- phyloseq::estimate_richness(
  full_meta, measures = c("Shannon", "Simpson", "Observed")) %>%
  as_tibble(rownames = "SampleID") %>%
  left_join(phyloseq::sample_data(physeq), by = 'SampleID')

alpha.df.l <- alpha.df %>%
  pivot_longer(cols = c("Shannon", "Simpson", "Observed"),
               names_to = 'Diversity_Measure', values_to = 'value')

alpha.box1 <- ggplot(alpha.df.l %>% filter(Diversity_Measure != 'Simpson')) +
  geom_boxplot(aes(y = value, fill = ACVD)) +
  ylab('Alpha Diversity') +
  scale_fill_manual(values = acvd_colors) +
  facet_wrap(~Diversity_Measure, scales = 'free') +
  my_gg_theme
alpha.box1

# alpha diversity stats
#' @examples
#' print(kruskal.test(data = alpha.df, Observed ~ ACVD))
#' print(kruskal.test(data = alpha.df, Shannon ~ ACVD))


# wilcox test
stats <- compare_means(data = alpha.df, formula = c(Observed, Shannon) ~ ACVD,
                       p.adjust.method = 'fdr')
print(stats)

g <- ggplot(alpha.df) +
  geom_point(aes(x = Observed, y = Shannon, color = ACVD), size = 3) +
  scale_color_manual(values = acvd_colors) +
  my_gg_theme
#' @examples
#' g
```


## 3.2 Beta diversity -- (physeq, metadata)

Beta diversity assesses differences in microbiome composition between different groups of samples. Bray-Curtis dissimilarity calculates differences using both presence and abundance information.

```{r beta, fig.cap="Figure 3.2: PCA of Beta-diversity (Bray-Curtis) colored by ACVD status. Ellipses represent 99% confidence interval for samples from each group."}
# bray curtis distance
physeq.beta <- phyloseq::ordinate(full_meta, method = 'PCoA', distance = 'bray', )
#scree_df(physeq.beta)


p <- physeq %>%
  tax_transform("identity", rank = "Species") %>%
  dist_calc(dist = "aitchison") %>%
  ord_calc("PCoA") %>%
  ord_plot(color = "ACVD", size = 2) +
  scale_color_manual(values = acvd_colors) +
  scale_fill_manual(values = acvd_colors) +
  #scale_colour_brewer(palette = "Dark2", aesthetics = c("fill", "colour"), name = "ACVD") +
  theme_bw() +
  ggside::geom_xsidedensity(aes(fill = ACVD), alpha = 0.5, show.legend = FALSE) +
  ggside::geom_ysidedensity(aes(fill = ACVD), alpha = 0.5, show.legend = FALSE) +
  ggside::theme_ggside_void()

scree_df()
scree_df <- tibble(
  comp_num = 1:length(physeq.beta$values$Eigenvalues),
  comp_nm = paste0("PCoA ", comp_num),
  var = physeq.beta$values$Eigenvalues,
  pct_var = var / sum(var) * 100,
  pct_cumvar = cumsum(pct_var)
)
d_80pct_var <- scree_df |>
  filter(pct_cumvar >= 80) |>
  slice(1) ## 42  :(
#d_math <- ide_math(pca.df[, -1]) %>% median(na.rm = TRUE) %>% round() #4
"Yeah, nah, round median ide_math of d=220 beta space was 4, lets use that"
d <- 4

pca.df <- data.frame(physeq.beta$vectors)
colnames(pca.df) <- paste0("PCoA", 1:ncol(pca.df))
pca.df <- tibble(pca.df[, 1:d],
                 SampleID = rownames(pca.df))

pca.df <- left_join(pca.df, metadata, by = 'SampleID') %>%
  mutate(bmi_bin = cut(BMI, breaks = c(0, 18, 25, 30, 35, 40, 50, 70),
                       include.lowest = TRUE, right = FALSE, dig.lab = 4))



# flat plots
#' @examples
#' pca.legend <- ggplot(pca.df,aes(color=Treatment, shape=Timepoint)) +
#'   geom_point()
#' pca.legend

pca.plot <- ggplot(pca.df) +
  geom_point(aes(x = PCoA1, y = PCoA2, color = ACVD), size = 2.5) +
  stat_ellipse(aes(x = PCoA1, y = PCoA2, color = ACVD),
               level = 0.99, linewidth = 1) +
  xlab(paste0("PC1 (", round(percent_explained[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(percent_explained[2], 1), "%)")) +
  scale_color_manual(values = acvd_colors) +
  my_gg_theme +
  labs(subtitle = 'Bray-Curtis Distance')
pca.plot
```

A permutational multivariate analysis of variance (PERMANOVA) is run to assess how differences in community composition can be attributed to different factors (ie disease state, age or gender). In this dataset community composition differs significantly by ACVD status and age, but not by gender. However, the amount of variance (R2) explained by these factors is very low: only 2% of variance is attributable to ACVD status, and <1% to age.

```{r beta_stats}
# testing beta diversity using Permanova
dist_matrix <- phyloseq::distance(full_meta, method = "bray")

set.seed("2507") #for reproducibility
beta_metadata <- data.frame(sample_data(full_meta))
permanova_result1 <- vegan::adonis2(dist_matrix ~ ACVD + Age + Gender,
                                    method = "bray", data = beta_metadata,
                                    by = "terms", parallel = 4 )
print(permanova_result1)

beta_dispersion1 <- vegan::betadisper(dist_matrix, beta_metadata$ACVD)
#print(beta_dispersion1)

anova_betadisper1 <- vegan::permutest(beta_dispersion1, 
                                      pairwise = FALSE, parallel = 4)
#print(anova_betadisper1)

tukeyhsb_betadisper1 <- TukeyHSD(beta_dispersion1)
#print(tukeyhsb_betadisper1)
```


## 3.3 Microbiome Composition Plots


### Core taxa by Family

Taxa filtered by 1% abundance and 20% prevalence and summarized at family level. Relative abundance is averaged by group (ACVD status).

```{r composition1, fig.cap="Figure 3.3: Relative abundance of core families (20% prevalence & 1% abundance) plotted by ACVD status."}
# remove unclassified taxa for plotting core families
physeq_nounclass <- subset_taxa(physeq, Family != 'UNCLASSIFIED')

# relative abundance - all taxa
physeq_nounclass_rel <- transform_sample_counts(
  physeq_nounclass, function(x) x / sum(x))

core <- aggregate_rare(physeq_nounclass_rel, level = "Family",
                       detection = 0.01, prevalence = 0.2)
core_families <- rownames(tax_table(core))
mycols <- microViz::distinct_palette(n = length(core_families),
                                     pal = 'greenArmytage')
mycols2 <- setNames(rev(mycols), core_families)

core.plot <- plot_composition(core, otu.sort = "abundance", average_by = 'ACVD') + 
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "ACVD",  y = "Relative abundance",
       subtitle = "Abundance = 1%, Prevalence = 20%", fill = 'Family') +
  my_gg_theme

c1<- core.plot + scale_fill_manual(values = mycols2)
ggplotly(c1)

#physeq %>%
#  ps_mutate(gender_acvd = interaction(Gender, ACVD)) %>%
#  phyloseq::merge_samples(group = "gender_acvd") %>%
#  comp_barplot(
#    tax_level = "Genus", n_taxa = 15,
#    sample_order = c('male.1','male.0','female.1','female.0'),
#    bar_width = 0.8
#  ) +
# coord_flip() + labs(x = NULL, y = NULL) +
#  theme(axis.text = element_text(size = 12),
#        legend.text = element_text(size=12),
#        legend.title = element_text(size=12,face='bold'))
```


## 3.4 Akkermansia {#akk}

```{r akk_abd, include=FALSE}
physeq_rel <- transform_sample_counts(physeq, function(x) x / sum(x))
physeq_smg_taxa <- as.data.frame(tax_table(physeq_rel)) %>%
  rownames_to_column(var = 'metaphlan_species') %>%
  select(metaphlan_species, Species)
abd_smg_rel <- as.data.frame(otu_table(physeq_rel)) %>%
  rownames_to_column(var = 'metaphlan_species') %>%
  left_join(physeq_smg_taxa, by = 'metaphlan_species') %>%
  select(-metaphlan_species) %>%
  column_to_rownames(var = 'Species')
df_rel <- t(abd_smg_rel)

abd_akks <- data.frame(t(df_rel)) %>%
  rownames_to_column(var = 'Species') %>%
  pivot_longer(cols = -c(Species), names_to = 'SampleID', values_to = 'abundance') %>%
  filter(str_detect(Species, "^Akkermansia")) %>%
  left_join(metadata, by = 'SampleID') %>%
  mutate(abd_perc = abundance * 100) %>%
  filter(Species %in% c("Akkermansia_muciniphila", "Akkermansia_massiliensis",
                        "Akkermansia_sp_N21116")) %>%
  mutate(Species = factor(
    replace(
      Species, Species == "Akkermansia_sp_N21116", "Akkermansia_guangxiensis"),
    levels = c("Akkermansia_muciniphila", "Akkermansia_massiliensis",
               "Akkermansia_guangxiensis")))

abd_akks.w <- abd_akks %>%
  pivot_wider(id_cols = c(SampleID, ACVD, TRIG, LDLC, HDLC, CHOL),
              names_from = Species, values_from = abd_perc)
#write.table(file='./outputs/acvd_abd_akks_w.tsv',abd_akks.w,quote = F,row.names = F,sep='\t')

akk_order <- c("Akkermansia_muciniphila", "Akkermansia_massiliensis", "Akkermansia_guangxiensis")
```

This table shows the prevalence and abundance of _Akkermansia_ species in each group (healthy and ACVD). Note that the mean and median relative abundance shown here are calculated including only individuals where the taxon is present (ie when this species is present in an individual, what is its abundance?). Note that comparing MEAN relative abundance between groups provides a different result to comparing MEDIAN relative abundance.

```{r akkprevabund}
# per age group
akk_abd_summ <- abd_akks %>%
  group_by(Species, ACVD) %>%
  #filter(abd_perc > 0) %>%
  summarize(mean_abd = mean(abd_perc),
            median_abd = median(abd_perc),
            .groups = 'drop') %>%
  mutate(SA = paste(Species, ACVD, sep = '-')) %>%
  select(SA, mean_abd, median_abd)

akk_abd_summ_no0 <- abd_akks %>%
  group_by(Species,ACVD) %>%
  filter(abd_perc > 0) %>%
  summarize(mean_abd_no0 = mean(abd_perc),
            median_abd_no0 = median(abd_perc),
            .groups = 'drop') %>%
  mutate(SA = paste(Species, ACVD, sep = '-')) %>%
  select(SA, mean_abd_no0, median_abd_no0)

akk <- left_join(akk_abd_summ, akk_abd_summ_no0, by = 'SA') %>%
  separate_wider_delim(SA, delim = '-',
                       names = c('Species', 'ACVD'), cols_remove = FALSE) %>%
  mutate(mean_abd_no0 = replace_na(mean_abd_no0, 0),
         median_abd_no0 = replace_na(median_abd_no0, 0)) %>%
  select(SA,mean_abd, mean_abd_no0, median_abd, median_abd_no0)

# prevalence
prevalence.est<-function(df, taxa.are.rows = T){
  df$abundance[df$abundance>0] <- 1
  samp <- length(unique(df$SampleID))
  summ <- df %>% group_by(Species, ACVD) %>%
    summarize(count = sum(abundance), .groups = 'drop')
  return(summ)
}

df.prev <- prevalence.est(abd_akks)

prev <- metadata %>%
  group_by(ACVD) %>%
  summarize(sample_count = n(), .groups = 'drop') %>%
  select(ACVD, sample_count)

summ <- df.prev %>%
  #filter(taxa %in% c('s__Akkermansia_massiliensis','s__Akkermansia_muciniphila')) %>%
  #group_by(Treatment,Timepoint) %>%
  #summarize(taxa_count = sum(count), .groups='drop') %>%
  left_join(prev, by = 'ACVD', multiple = 'first') %>%
  mutate(prevalence = round(count / sample_count * 100, digits = 2)) %>%
  mutate(SA = paste(Species, ACVD, sep = '-')) %>%
  left_join(akk, by = 'SA') %>%
  mutate(mean_abd_rounded = round(mean_abd, 2),
         mean_abd_no0_rounded = round(mean_abd_no0, 2),
         median_abd_rounded = round(median_abd, 2),
         median_abd_no0_rounded = round(median_abd_no0, 2)) %>%
  mutate(prev_perc_n = paste0(round(prevalence, 0), '% (', count, ')')) %>%
  select(Species, ACVD,
         'N' = sample_count,
         'prevalence(N)' = prev_perc_n,
         'mean_abundance' = mean_abd_no0_rounded,
         'median_abundance' = median_abd_no0_rounded)


DT::datatable(
  summ %>%
    filter(Species %in% c('Akkermansia_muciniphila','Akkermansia_massiliensis','Akkermansia_guangxiensis')),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; color:black; font-size:150%;',
    'Akkermansia prevalence and abundance by group and visit'),
  rownames = FALSE,
  width = '100%',
  options = list(
    scrollX = '500px',
    scrollY = '300px',
    paging = FALSE,
    info = FALSE,
    initComplete = JS("function(settings, json) {$(this.api().table().header()).css({'font-size' : '14px'});}")
  )
) %>%
  formatStyle(columns = colnames(.$x$data), `font-size` = '14px')
```

Here we are plotting the relative abundance of _Akkermansia_ species broken down by ACVD status: 0 = healthy, 1 = ACVD. Note that the y-axis is on a log scale for visibility.

With this plot we can see how a comparison of the MEDIANS tells us that the abundance of both _A. muciniphila_ and _A. massiliensis_ are lower in people with ACVD, when compared to healthy individuals. We can also see how my initial comparison of the MEANS was influenced by a few individuals who have abnormally high (>10%) abundance of _A. muciniphila_.

<a id="akk-box"></a>

```{r akk, fig.cap="Figure 3.4.1: Akkermansia abundance by species and ACVD status. Note that the y-axis is on a log scale for visibility."}
akk_box <- ggplot(abd_akks %>% filter(Species != 'Akkermansia_biwaensis' & abd_perc > 0),
                  aes(x = Species, y = abd_perc)) +
  geom_boxplot(aes(fill = ACVD), outliers = F) +
  geom_point(aes(group = ACVD), position = position_dodge(width = 0.75),,
             color = 'grey27', alpha = 0.5) +
  my_gg_theme +
  scale_y_log10() +
  scale_fill_manual(values = acvd_colors) +
  labs(y = 'Relative abundance (%)')
akk_box
```

Relative abundance of each Akkermansia sp. scaled to the total Akkermansia abundance plotted only for samples with at least 1 Akkermansia species present. Majority of samples contain only a single species of Akkermansia. Where multiple species are present, one species is usually very dominant, with only a small fraction of a secondary species.

```{r akk2, fig.width=8, fig.cap="Figure 3.4.2: Relative abundance of Akkermansia sp. in only Akkermansia+ individuals."}
#library(microViz)
akk_order <- c('Akkermansia_muciniphila', 'Akkermansia_massiliensis',
               'Akkermansia_guangxiensis')
akk_order2 <- c('Akkermansia_muciniphila', 'Akkermansia_massiliensis',
                'Akkermansia_sp_N21116')

no_akks <- abd_akks.w %>%
  mutate(across(all_of(akk_order), as.numeric)) %>%
  filter(if_all(akk_order, ~ .x == 0)) %>%
  pull(SampleID)

akk_samp <- abd_akks.w %>%
  mutate(across(all_of(akk_order), as.numeric)) %>%
  filter(!if_all(akk_order, ~ .x == 0)) %>%
  pull(SampleID)

psq <- physeq %>%
  #tax_mutate(Species = NULL) %>%
  #tax_filter(min_prevalence = 5) %>%
  tax_fix()

akk <- tax_select(psq, tax_list = 'Akkermansia') %>%
  ps_filter(!SampleID %in% no_akks) %>%
  tax_select(tax_list = 'Akkermansia_biwaensis', deselect = TRUE)

cols <- distinct_palette(n = 2, add = NA)
cols <- acvd_colors
names(cols) <- unique(samdat_tbl(akk)$ACVD)

htmp <- akk %>% tax_agg(rank = 'Species') %>%
  tax_reorder(tax_order = akk_order2) %>%
  tax_transform("compositional", rank = "Species") %>%
  comp_heatmap(
    taxa = akk_order2,
    samples = akk_samp,
    taxon_renamer = function(x) stringr::str_replace(
      stringr::str_replace(x, "Akkermansia_","A. "),"sp_N21116","guangxiensis"),
    #colors = heat_palette(rev=TRUE, breaks=8),
    colors = heat_palette(palette = colorRampPalette(c('white','steelblue','red4'))(10)),
    cluster_rows = FALSE,
    tax_seriation = 'Identity',
    #tax_anno = taxAnnotation(Prev. = anno_tax_prev(bar_width = 0.3, size=grid::unit(1,'cm'))),
    sample_anno = sampleAnnotation(ACVD = anno_sample_cat('ACVD', col=cols)),
    row_names_gp = grid::gpar(fontsize = 10),
    heatmap_legend_param = list(at = 0:5 / 5,
                                direction = "horizontal", title_position = "leftcenter",
                                legend_width = grid::unit(4, "cm"),
                                grid_height = grid::unit(5, "mm")))
#png(file='~/Documents/WORK/2026.01.30_ACVD_metagenomics/figures/akkermansia_heatmap.png',
#    width=10,height=4,units='in',res=600)

ComplexHeatmap::draw(
  object = htmp,
  heatmap_legend_side = 'bottom',
  annotation_legend_list = attr(htmp, "AnnoLegends"), merge_legends = TRUE)
#dev.off()
```



# 8 Save artifacts for presentation

```{r}
alpha.df.l
```


# 9 Session Info

```{r, results = "asis"}
tictoc::toc()
cat(.context)
```

```{r}
sessionInfo()
```
